Pervasive duplication of tumor suppressor genes preceded parallel evolution of large bodied Atlantogenatans

### 2020-05-28 paper_PLOS/paper_PLOS_draft.tex
Figure 1 corrected

% Template for PLoS
% Version 3.5 March 2018
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended
% to minimize problems and delays during our production
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% Once your paper is accepted for publication,
% PLEASE REMOVE ALL TRACKED CHANGES in this file
% and leave only the final text of your manuscript.
% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
%
%
% There are no restrictions on package use within the LaTeX files except that
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file.
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission.
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - spacing/line breaks within cells to alter layout or alignment
% - do not nest tabular environments (no tabular environments within tabular environments)
% - no graphics or colored text (cell background color/shading OK)
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
%
% For inline equations, please be sure to include all portions of an equation in the math environment.
%
% Do not include text that is not math in the math environment.
%
% Please add line breaks to long display equations when possible in order to fit size of the column.
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% When adding superscript or subscripts outside of brackets/braces, please group using {}.
%
% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% Use Unicode characters when possible
\usepackage[utf8x]{inputenc}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% cite package, to clean up citations in the main text. Do not remove.
% \usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% color can be used to apply background shading to table cells only
\usepackage[table]{xcolor}

% array package and thick rules for tables
\usepackage{array}

% create "+" rule type for thick vertical lines
\newcolumntype{+}{!{\vrule width 2pt}}

% create \thickcline for thick horizontal lines of variable length
\newlength\savedwidth
\newcommand\thickcline[1]{%
  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
  \cline{#1}%
  \noalign{\vskip\arrayrulewidth}%
  \noalign{\global\arrayrulewidth\savedwidth}%
}

% \thickhline command for thick horizontal lines that span the table
\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
\hline
\noalign{\global\arrayrulewidth\savedwidth}}

% Remove comment for double spacing
%\usepackage{setspace}
%\doublespacing

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

% Use the PLoS provided BiBTeX style
% \bibliographystyle{plos2015}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
%\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
%\setlength{\headheight}{27.023pt}
%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\today}

%% Include all macros below

\newcommand{\lorem}{{\bf LOREM}}
\newcommand{\ipsum}{{\bf IPSUM}}

\usepackage{longtable}
\usepackage{booktabs}
\usepackage{geometry}
\usepackage{pdflscape}

\usepackage{forarray}
\usepackage{xstring}
\newcommand{\getIndex}[2]{
  \ForEach{,}{\IfEq{#1}{\thislevelitem}{\number\thislevelcount\ExitForEach}{}}{#2}
}

\setcounter{secnumdepth}{0}

\newcommand{\getAff}[1]{
  \getIndex{#1}{University of Chicago,University at Buffalo, SUNY}
}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\begin{document}
\vspace*{0.2in}

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Pervasive duplication of tumor suppressor genes preceded parallel
evolution of large bodied Atlantogenatans} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Juan Manuel Vazquez\textsuperscript{\getAff{University of Chicago}}\textsuperscript{*},
Vincent J Lynch\textsuperscript{\getAff{University at Buffalo, SUNY}}\\
\bigskip
\textbf{\getAff{University of Chicago}}Department of Human Genetics, 920 East 58th St, Chicago, IL, 60637\\
\textbf{\getAff{University at Buffalo, SUNY}}Department of Biological Sciences, 551 Cooke Hall, Buffalo NY, 14260\\
\bigskip
* Corresponding author: juanvazquez@uchicago.edu\\
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Cancer is an intrinsic disease of multicellular organisms. Within a
species, the size of an animal, - correlated with the individual's
number of cells - and its lifespan - correlated with increasing cellular
damage over time - positively correlate with the risk any individual has
to form tumors. Between species, however, we do not observe any
correlation between size, lifespan, and cancer, a phenomenon that
referred to as \emph{Peto's Paradox}. Elephants are a particularly
interesting member of this class of paradoxical animals, since they are
a set of large species deeply nested in a clade of smaller species,
indicating a recent gain of size. Recent work has identified several
individual cases of gene duplicates contributing to the increased cancer
resistance of elephants, which suggests that duplication of tumor
suppressor genes may play a more general role in mediating Peto's
Paradox by increasing cancer resistance in large, long-lived species. By
using a Reciprocal Best-Hit BLAT search approach, we investigated copy
numbers of all protein-coding genes in \emph{Atlantogenatan} genomes to
see if there is any correlation between the copy number of duplicates
and changes body size along the phylogenetic tree. From an initial set
of 18,011 protein-coding genes in hg38, we identified a median of 13,880
genes in \emph{Atlantogenatan} genomes, of which a median of 940 genes
are duplicated. We find that, just as body size fluctuates throughout
\emph{Atlantogenata}, genes involved in tumor suppressor pathways are
also duplicated throughout the phylogenetic tree. Extant species of
elephants, however, show active transcription of both canonical and
duplicated copies of tummor suppressors that duplicated prior to and
during their sudden increase in body size, suggesting that the
duplication of tumor suppressor genes facilitates the evolution of
increased body size by compensating for the increased cancer risk.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step.
% Author Summary not valid for PLOS ONE submissions.
\section*{Author summary}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget
porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut
dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id
velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus
finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium
diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut
ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit
pulvinar, volutpat risus.

\linenumbers

% Use "Eq" instead of "Equation" for equation citations.
)

@@

One of the major constraints on the evolution of large body sizes in
animals is an increased risk of developing cancer. If all cells in all
organisms have a similar risk of malignant transformation and equivalent
cancer suppression mechanisms, organism with many cells should have a
higher prevalence of cancer than organisms with fewer cells. Consistent
with this expectation there is a strong positive correlation between
body size and cancer incidence within species, for example, human cancer
incidence increases with increasing adult height {[}1,2{]} and cancer
incidence is positively correlated with body size in dogs
{[}{\textbf{???}},3{]}. There is no correlation, however, between body
size and cancer risk between species. This lack of correlation is often
referred to as `Peto's Paradox' {[}4--6{]}. While it is clear that a
resolution to Peto's Paradox must involve the evolution of enhanced
cancer protection alongside increases in body size and lifespan, the
specific genetic, molecular, and cellular mechanisms that underlie this
resistance have proven elusive. {[}7--11{]}.

Among the challenges for discovering how animals evolved enhanced cancer
protection mechanisms is identifying lineages in which large bodied
species are nested within species with small body sizes. Afrotherian
mammals are generally small-bodied, similarly to the prediced common
ancestor of Eutherian mammals. For example, maximum adult weights are
\textasciitilde{}70g in golden moles, \textasciitilde{}120g in tenrecs,
\textasciitilde{}170g in elephant shrews, \textasciitilde{}3kg in
hyraxes, and 60kg in aardvarks {[}12{]}. However, while these extant
species are relatively small, the fossil evidence demonstrates that
their ancestral lineages reached enormous sizes. For example, while
extant hyraxes are relatively small, the extinct Titanohyrax is
estimated to have weighted up to \textasciitilde{}1300kg {[}13{]}. The
largest members of Afrotheria, too, are dwarfed by the size of their
recent ancestors: extant cows manatees are large bodied
(\textasciitilde{}322-480kg) but are relatively small compared to the
extinct Stellar's sea cow which is estimated to have weight 8000-10000kg
{[}14{]}. Similarly African (4,800kg) and Asian elephants (3,200kg) are
the largest living elephant species, but are dwarfed by the truly
gigantic extinct Proboscideans such as Deinotherium
(\textasciitilde{}132,000kg), Mammut borsoni (110,000kg), and the Asian
straight-tusked elephant (\textasciitilde{}220,000kg), the largest known
land mammal {[}15{]}. Remarkably these large-bodied Afrotherian lineages
are nested within small bodied species (Fig. 1) {[}16--19{]}, indicating
that gigantism independently evolved in hyraxes, sea cows, and elephants
(Paenungulates). Thus, Paenungulates are an excellent model system in
which to explore the mechanisms that underlie the evolution of large
body sizes and augmented cancer resistance.

Although many mechanisms can potentially resolve Peto's paradox, the
most parsimonious route to enhanced cancer resistance is likely through
an increased copy number of tumor suppressors. Such an example has been
seen in the case of candidate genes such as \emph{TP53} and \emph{LIF}
{[}11,20,21{]} as well as in studies involving a limited set of
candidate genes {[}22,23{]}. As these studies focus on \emph{a priori}
gene sets, however, it remains unknown whether this is a general,
genome-wide trend in Afrotherian genomes; and whether such a general
trend is associated with the recent increases in body size -- and
therefore expected cancer risk -- in these species.

Here, we trace the evolution of body mass and gene copy number variation
in Afrotherians in order to investigate whether gene duplications are
enriched in large, long-lived species for genes involved in known tumor
suppression pathways. Our estimates of the evolution of body mass,
similarly to previous studies {[}16--19{]}, show that large body masses
evolved in a step-wise manner, with major increases in body mass in the
Pseudoungulata (17kg), Paenungulata (25kg), Tethytheria (296kg), and
Proboscidea (4,100kg) stem-lineages. Furthermore, we see that the
ancestral body size increases in Hydracoidia and Sirenia were
independent events. To study the evolution of gene copy number, we used
a genome-wide Reciprocal Best BLAT Hit (RBBH) method to identify gene
duplications in Afrotherian genomes, and used parsimony to infer the
lineages in which those duplications occurred. We found gene
duplications in lineages with increased body mass were enriched in
functions related to tumor suppression, including regulation of the cell
cycle, DNA damage repair, and regulation of apoptosis. These data
suggest that duplication of tumor suppressors played a role in the
evolution of large, long-lived in Afrotherians.

@@

We built a time-calibrated supertree of Eutherian mammals by combining
the time-calibrated molecular phylogeny of Bininda-Emonds \emph{et al.}
{[}24{]} with the time-calibrated total evidence Afrotherian phylogeny
from Puttick and Thomas {[}{\textbf{???}}{]}. While the Bininda-Emonds
\emph{et al.} {[}24{]} phylogeny includes 1,679 species, only 34 are
Afrotherian, and no fossil data are included. The inclusion of fossil
data from extinct species is essential to ensure that ancestral state
reconstructions of body mass are not biased by only including extant
species. This can lead to inaccurate reconstructions, for example, if
lineages convergently evolved large body masses from a small bodied
ancestor. In contrast, the total evidence Afrotherian phylogeny of
Puttick and Thomas {[}19{]} includes 77 extant species and fossil data
from 39 extinct species. Therefore we replaced the Afrotherian clade in
the Bininda-Emonds \emph{et al.} {[}24{]} phylogeny with the Afrotherian
phylogeny of Puttick and Thomas {[}19{]} using Mesquite. Next, we
jointly estimated rates of body mass evolution and reconstructed
ancestral states using a generalization of the Brownian motion model
that relaxes assumptions of neutrality and gradualism by considering
increments to evolving characters to be drawn from a heavy-tailed stable
distribution (the ``Stable Model'') {[}25{]}. The stable model allows
for occasional large jumps in traits and has previously been shown to
out-perform other models of body mass evolution, including standard
Brownian motion models, Ornstein--Uhlenbeck models, early burst maximum
likelihood models, and heterogeneous multi-rate models {[}25{]}.

@@

\textbf{Reciprocal Best-Hit BLAT:} We developed a reciprocal best hit
BLAT (RBHB) pipeline to quickly identify homologs and estimate gene copy
numbers (\textbf{Figure 1A}). The Reciprocal Best Hit (RBH) search
strategy is conceptually straightforward: 1) Given a gene of interest
\(G_A\) in a query genome \(A\), one searches a target genome \(B\) for
all possible matches to \(G_A\); 2) For each of these hits, one then
performs the reciprocal search in the original query genome to identify
the highest-scoring hit; 3) A hit in genome \(B\) is defined as a
homolog of gene \(G_A\) if and only if the original gene \(G_A\) is the
top reciprocal search hit in genome \(A\). We selected BLAT {[}26{]} as
our algorithm of choice, as this algorithm is sensitive to highly
simliar (\textgreater{}90\% identity) sequences, thus identifying the
highest-confidence homologs while minimizing many-to-one mapping
problems when searching for multiple genes. RBH performs similar to
other more complex methods of orthology prediction, and is particularly
good at identifying incomplete genes that may be fragmented in low
quality/poor assembled regions of the genome {[}{\textbf{???}},27{]}.

@@

\textbf{RecSearch Pipeline} We created a custom Python pipeline for
automating RBHB searches between a single reference genome and multiple
target genomes using a list of query sequences from the reference
genome. For the query sequences in our search, we used the hg38 Proteome
provided by UniProt {[}28{]}, which is a comprehensive set of protein
sequences curated from a combination of predicted and validated protein
sequences generated by the UniProt Consortium. In order to refine our
search, we omitted protein sequences originating from long, noncoding
RNA loci (e.g.~LINC genes); poorly-studied genes from predicted open
reading frames (C-ORFs); and sequences with highly repetitive sequences
such as zinc fingers, protocadherins, and transposon-containing genes,
as these were prone to high levels of false positive hits. After
filtering out problematic protein queries, we then used our pipeline
(Figure 1A) to search for all copies of our 18011 query genes in
publicly available Afrotherian genomes, including African savannah
elephant (\emph{Loxodonta africana}: loxAfr3, loxAfr4, loxAfrC), African
forest elephant (\emph{Loxodonta cyclotis}: loxCycF), Asian Elephant
(\emph{Elephas maximus}: eleMaxD), Woolly Mammoth (\emph{Mammuthus
primigenius}: mamPriV), Colombian mammoth (\emph{Mammuthus columbi}:
mamColU), American mastodon (\emph{Mammut americanum}: mamAmeI), Rock
Hyrax (\emph{Procavia capensis}: proCap1, proCap2, proCap2\_HiC), West
Indian Manatee (\emph{Trichechus manatus latirostris}: triManLat1,
triManLat1\_HiC), Aardvark (\emph{Orycteropus afer}: oryAfe1,
oryAfe1\_HiC), Lesser Hedgehog Tenrec (\emph{Echinops telfairi}:
echTel2), Nine-banded armadillo (\emph{Dasypus novemcinctus}: dasNov3),
Hoffman's two-toed sloth (\emph{Choloepus hoffmannii}: choHof1, choHof2,
choHof2\_HiC), Cape golden mole (\emph{Chrysochloris asiatica}:
chrAsi1), and Cape elephant shrew (\emph{Elephantulus edwardii}:
eleEdw1). For many of these species, we covered multiple assemblies in
order to test the effects of assembly size and quality on our hits.

\textbf{Query protein criteria} To assemble our query list, we first
removed all unnamed genes from UP000005640. Next, we excluded genes from
downstream analyses for which assignment of homology was uncertain,
including uncharacterized ORFs (991), LOC (63), HLA genes (402),
replication dependent histones (72), odorant receptors (499), ribosomal
proteins (410), zinc finger transcription factors (1983), viral and
repetitive-element-associated proteins (82) and any protein described as
either ``Uncharacterized,'' ``Putative,'' or ``Fragment'' by UniProt in
UP000005640 (30724), leaving us with a final set of 37582 query protein
sequences, corresponding to 18011 genes.

\textbf{Duplication gene inclusion criteria} In order to condense
transcript-level hits into single gene loci, and to resolve many-to-one
genome mappings, we removed exons where transcripts from different genes
overlapped, and merged overlapping transcripts of the same gene into a
single gene locus call. The resulting gene-level copy number table was
then combined with the maximum ECNC values observed for each gene in
order to call gene duplications. We called a gene duplicated if its copy
number was two or more, and if the maximum ECNC value of all the gene
transcripts searched was 1.5 or greater; previous studies have shown
that incomplete duplications can encode functional genes, therefore
partial gene duplications were included provided they passed additional
inclusion criteria. The ECNC cut off of 1.5 was selected empirically, as
this value minimized the number of false positives seen in a test set of
genes and genomes. The results of our initial search are summarized in
Figure 1B. Overall, we identified 13880 genes across all species, or
77.1\% of our starting query genes.

>>>>

\documentclass[]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
\usepackage[hyphens]{url}

\usepackage{lineno} % add
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{graphicx}
\usepackage{booktabs} % book-quality tables
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Pervasive duplication of tumor suppressor genes preceded parallel evolution of large bodied Atlantogenatans},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}

% Pandoc header
\usepackage{longtable}
\usepackage{booktabs}
\usepackage{geometry}
\usepackage{pdflscape}

\begin{document}
\begin{frontmatter}

  \title{Pervasive duplication of tumor suppressor genes preceded parallel
evolution of large bodied Atlantogenatans}
    \author[University of Chicago]{Juan Manuel Vazquez}

    \author[SUNY Buffalo]{Vincent J Lynch}

      \address[University of Chicago]{Department of Human Genetics, 920 East 58th St, Chicago, IL, 60637}
    \address[SUNY Buffalo]{Department of Biological Sciences, 551 Cooke Hall, Buffalo NY, 14260}

  \begin{abstract}
  Cancer is an intrinsic disease of multicellular organisms. Within a
  species, the size of an animal, - correlated with the individual's
  number of cells - and its lifespan - correlated with increasing cellular
  damage over time - positively correlate with the risk any individual has
  to form tumors. Between species, however, we do not observe any
  correlation between size, lifespan, and cancer, a phenomenon that
  referred to as \emph{Peto's Paradox}. Elephants are a particularly
  interesting member of this class of paradoxical animals, since they are
  a set of large species deeply nested in a clade of smaller species,
  indicating a recent gain of size. Recent work has identified several
  individual cases of gene duplicates contributing to the increased cancer
  resistance of elephants, which suggests that duplication of tumor
  suppressor genes may play a more general role in mediating Peto's
  Paradox by increasing cancer resistance in large, long-lived species. By
  using a Reciprocal Best-Hit BLAT search approach, we investigated copy
  numbers of all protein-coding genes in \emph{Atlantogenatan} genomes to
  see if there is any correlation between the copy number of duplicates
  and changes body size along the phylogenetic tree. From an initial set
  of 18,011 protein-coding genes in hg38, we identified a median of 13,880
  genes in \emph{Atlantogenatan} genomes, of which a median of 940 genes
  are duplicated. We find that, just as body size fluctuates throughout
  \emph{Atlantogenata}, genes involved in tumor suppressor pathways are
  also duplicated throughout the phylogenetic tree. Extant species of
  elephants, however, show active transcription of both canonical and
  duplicated copies of tummor suppressors that duplicated prior to and
  during their sudden increase in body size, suggesting that the
  duplication of tumor suppressor genes facilitates the evolution of
  increased body size by compensating for the increased cancer risk.
  \end{abstract}

 \end{frontmatter}

)

@@

One of the major constraints on the evolution of large body sizes in
animals is an increased risk of developing cancer. If all cells in all
organisms have a similar risk of malignant transformation and equivalent
cancer suppression mechanisms, organism with many cells should have a
higher prevalence of cancer than organisms with fewer cells. Consistent
with this expectation there is a strong positive correlation between
body size and cancer incidence within species, for example, human cancer
incidence increases with increasing adult height {[}1,2{]} and cancer
incidence is positively correlated with body size in dogs {[}3,4{]}.
There is no correlation, however, between body size and cancer risk
between species. This lack of correlation is often referred to as
`Peto's Paradox' {[}5--7{]}. While it is clear that a resolution to
Peto's Paradox must involve the evolution of enhanced cancer protection
alongside increases in body size and lifespan, the specific genetic,
molecular, and cellular mechanisms that underlie this resistance have
proven elusive. {[}8--12{]}.

Among the challenges for discovering how animals evolved enhanced cancer
protection mechanisms is identifying lineages in which large bodied
species are nested within species with small body sizes. Afrotherian
mammals are generally small-bodied, similarly to the prediced common
ancestor of Eutherian mammals. For example, maximum adult weights are
\textasciitilde{}70g in golden moles, \textasciitilde{}120g in tenrecs,
\textasciitilde{}170g in elephant shrews, \textasciitilde{}3kg in
hyraxes, and 60kg in aardvarks {[}13{]}. However, while these extant
species are relatively small, the fossil evidence demonstrates that
their ancestral lineages reached enormous sizes. For example, while
extant hyraxes are relatively small, the extinct Titanohyrax is
estimated to have weighted up to \textasciitilde{}1300kg {[}14{]}. The
largest members of Afrotheria, too, are dwarfed by the size of their
recent ancestors: extant cows manatees are large bodied
(\textasciitilde{}322-480kg) but are relatively small compared to the
extinct Stellar's sea cow which is estimated to have weight 8000-10000kg
{[}15{]}. Similarly African (4,800kg) and Asian elephants (3,200kg) are
the largest living elephant species, but are dwarfed by the truly
gigantic extinct Proboscideans such as Deinotherium
(\textasciitilde{}132,000kg), Mammut borsoni (110,000kg), and the Asian
straight-tusked elephant (\textasciitilde{}220,000kg), the largest known
land mammal {[}16{]}. These large-bodied Afrotherian lineages are nested
within small bodied species (Fig. 1) {[}17--20{]}, indicating that
gigantism independently evolved in hyraxes, sea cows, and elephants
(Paenungulates). Thus, Paenungulates are an excellent model system in
which to explore the mechanisms that underlie the evolution of large
body sizes and augmented cancer resistance.

Although many mechanisms can potentially resolve Peto's paradox, among
the most parsimonious routes to enhanced cancer resistance is through an
increased copy number of tumor suppressors. Such an example has been
seen in the case of candidate genes such as \emph{TP53} and \emph{LIF}
{[}12,21,22{]} as well as in studies involving a limited set of
candidate genes {[}23,24{]}. As these studies focus on \emph{a priori}
gene sets, however, it remains unknown whether this is a general,
genome-wide trend in Afrotherian genomes; and whether such a general
trend is associated with the recent increases in body size -- and
therefore expected cancer risk -- in these species.

Here, we trace the evolution of body mass and gene copy number variation
in Afrotherians in order to investigate whether gene duplications are
enriched in large, long-lived species for genes involved in known tumor
suppression pathways. Our estimates of the evolution of body mass,
similarly to previous studies {[}17--20{]}, show that large body masses
evolved in a step-wise manner, with major increases in body mass in the
Pseudoungulata (17kg), Paenungulata (25kg), Tethytheria (296kg), and
Proboscidea (4,100kg) stem-lineages. Furthermore, we see that the
ancestral body size increases in Hydracoidia and Sirenia were
independent events. To study the evolution of gene copy number, we used
a genome-wide Reciprocal Best BLAT Hit (RBBH) method to identify gene
duplications in Afrotherian genomes, and used maximum likelihood
(treating copy number as a discrete trait) to infer the lineages in
which those duplications occurred. We found gene duplications in
lineages with increased body mass were enriched in functions related to
tumor suppression, including regulation of the cell cycle, DNA damage
repair, and regulation of apoptosis. These data suggest that duplication
of tumor suppressors played a role in the evolution of large, long-lived
in Afrotherians.

@@

We built a time-calibrated supertree of Eutherian mammals by combining
the time-calibrated molecular phylogeny of Bininda-Emonds \emph{et al.}
{[}25{]} with the time-calibrated total evidence Afrotherian phylogeny
from Puttick and Thomas {[}20{]}. While the Bininda-Emonds \emph{et al.}
{[}25{]} phylogeny includes 1,679 species, only 34 are Afrotherian, and
no fossil data are included. The inclusion of fossil data from extinct
species is essential to ensure that ancestral state reconstructions of
body mass are not biased by only including extant species. This can lead
to inaccurate reconstructions, for example, if lineages convergently
evolved large body masses from a small bodied ancestor. In contrast, the
total evidence Afrotherian phylogeny of Puttick and Thomas {[}20{]}
includes 77 extant species and fossil data from 39 extinct species.
Therefore we replaced the Afrotherian clade in the Bininda-Emonds
\emph{et al.} {[}25{]} phylogeny with the Afrotherian phylogeny of
Puttick and Thomas {[}20{]} using Mesquite. Next, we jointly estimated
rates of body mass evolution and reconstructed ancestral states using a
generalization of the Brownian motion model that relaxes assumptions of
neutrality and gradualism by considering increments to evolving
characters to be drawn from a heavy-tailed stable distribution (the
``Stable Model'') {[}26{]}. The stable model allows for occasional large
jumps in traits and has previously been shown to out-perform other
models of body mass evolution, including standard Brownian motion
models, Ornstein--Uhlenbeck models, early burst maximum likelihood
models, and heterogeneous multi-rate models {[}26{]}.

@@

\textbf{Reciprocal Best-Hit BLAT:} We developed a reciprocal best hit
BLAT (RBHB) pipeline to identify putative homologs and estimate gene
copy numbers across species (\textbf{Figure 1A}). The Reciprocal Best
Hit (RBH) search strategy is conceptually straightforward: 1) Given a
gene of interest \(G_A\) in a query genome \(A\), one searches a target
genome \(B\) for all possible matches to \(G_A\); 2) For each of these
hits, one then performs the reciprocal search in the original query
genome to identify the highest-scoring hit; 3) A hit in genome \(B\) is
defined as a homolog of gene \(G_A\) if and only if the original gene
\(G_A\) is the top reciprocal search hit in genome \(A\). We selected
BLAT {[}27{]} as our algorithm of choice, as this algorithm is sensitive
to highly simliar (\textgreater{}90\% identity) sequences, thus
identifying the highest-confidence homologs while minimizing many-to-one
mapping problems when searching for multiple genes. RBH performs similar
to other more complex methods of orthology prediction, and is
particularly good at identifying incomplete genes that may be fragmented
in low quality/poor assembled regions of the genome {[}28,29{]}.

@@

\textbf{RecSearch Pipeline:} We created a custom Python pipeline for
automating RBHB searches between a single reference genome and multiple
target genomes using a list of query sequences from the reference
genome. For the query sequences in our search, we used the hg38 Proteome
provided by UniProt {[}30{]}, which is a comprehensive set of protein
sequences curated from a combination of predicted and validated protein
sequences generated by the UniProt Consortium. In order to refine our
search, we omitted protein sequences originating from long, noncoding
RNA loci (e.g.~LINC genes); poorly-studied genes from predicted open
reading frames (C-ORFs); and sequences with highly repetitive sequences
such as zinc fingers, protocadherins, and transposon-containing genes,
as these were prone to high levels of false positive hits.

After filtering out problematic protein queries (see ``Query gene
inclusion criteria''), we then used our pipeline (Figure 1A) to search
for all copies of our 18011 query genes in publicly available
Afrotherian genomes, including African savannah elephant
(\emph{Loxodonta africana}: loxAfr3, loxAfr4, loxAfrC), African forest
elephant (\emph{Loxodonta cyclotis}: loxCycF), Asian Elephant
(\emph{Elephas maximus}: eleMaxD), Woolly Mammoth (\emph{Mammuthus
primigenius}: mamPriV), Colombian mammoth (\emph{Mammuthus columbi}:
mamColU), American mastodon (\emph{Mammut americanum}: mamAmeI), Rock
Hyrax (\emph{Procavia capensis}: proCap1, proCap2, proCap2\_HiC), West
Indian Manatee (\emph{Trichechus manatus latirostris}: triManLat1,
triManLat1\_HiC), Aardvark (\emph{Orycteropus afer}: oryAfe1,
oryAfe1\_HiC), Lesser Hedgehog Tenrec (\emph{Echinops telfairi}:
echTel2), Nine-banded armadillo (\emph{Dasypus novemcinctus}: dasNov3),
Hoffman's two-toed sloth (\emph{Choloepus hoffmannii}: choHof1, choHof2,
choHof2\_HiC), Cape golden mole (\emph{Chrysochloris asiatica}:
chrAsi1), and Cape elephant shrew (\emph{Elephantulus edwardii}:
eleEdw1). For many of these species, we covered multiple assemblies in
order to test the effects of assembly size and quality on our hits.

\textbf{Query gene inclusion criteria:} To assemble our query list, we
first removed all unnamed genes from UP000005640. Next, we excluded
genes from downstream analyses for which assignment of homology was
uncertain, including uncharacterized ORFs (991), LOC (63), HLA genes
(402), replication dependent histones (72), odorant receptors (499),
ribosomal proteins (410), zinc finger transcription factors (1983),
viral and repetitive-element-associated proteins (82) and any protein
described as either ``Uncharacterized,'' ``Putative,'' or ``Fragment''
by UniProt in UP000005640 (30724), leaving us with a final set of 37582
query protein sequences, corresponding to 18011 genes.

\textbf{Duplication gene inclusion criteria:} In order to condense
transcript-level hits into single gene loci, and to resolve many-to-one
genome mappings, we removed exons where transcripts from different genes
overlapped, and merged overlapping transcripts of the same gene into a
single gene locus call. The resulting gene-level copy number table was
then combined with the maximum ECNC values observed for each gene in
order to call gene duplications. We called a gene duplicated if its copy
number was two or more, and if the maximum ECNC value of all the gene
transcripts searched was 1.5 or greater; previous studies have shown
that incomplete duplications can encode functional genes {[}12,22{]},
therefore partial gene duplications were included provided they passed
additional inclusion criteria. The ECNC cut off of 1.5 was selected
empirically, as this value minimized the number of false positives seen
in a test set of genes and genomes. The results of our initial search
are summarized in Figure 1B. Overall, we identified 13880 genes across
all species, or 77.1\% of our starting query genes.
